An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-04-30 EconGrowth-R-Functions.Rnw
Partial residual plots for U now include only US, UK, and JP, the countries for which U data is available.

# Statistical significance levels. We'll work with 95% CIs
ciLevel <- 0.95
ciHalfLevel <- ciLevel + (1.0-ciLevel)/2.0
# List of countries
countryAbbrevs <- c(US="US", UK="UK", JP="JP", CN="CN", ZA="ZA", SA="SA", IR="IR", TZ="TZ", ZM="ZM")
countryAbbrevsAlph <- c(CN="CN", IR="IR", JP="JP", SA="SA", TZ="TZ", UK="UK", US="US", ZA="ZA", ZM="ZM")
countryAbbrevsU <- c(US="US", UK="UK", JP="JP") #Only these countries have useful work data
countryAbbrevsAlphU <- c(JP="JP", UK="UK", US="US")
countryNamesAlph <- c(CN="China", IR="Iran", JP="Japan", SA="Saudi Arabia", TZ="Tanzania", UK="United Kingdom", US="USA", ZA="South Africa", ZM="Zambia") #In alphabetical order.
yLimitsForGDPGraphs <- list(c(1,10), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4)) # Alph order
energyTypes <- c(Q="Q", X="X", U="U") # List of energy types
factors <- c(K="K", L="L", Q="Q", X="X", U="U") # List of factors of production
########### Several global parameters for graphs. Set here and use below to ensure consistent appearance of graphs.
# Set the order for presenting countries in 3x3 lattice graphs. Default is alphabetical. 
# "1" means first alphabetically.
countryOrderForGraphs <- c(7,6,3,1,8,5,4,2,9) # Sets the order as US, UK, JP, CN, ZA, TZ, SA, IR, ZM.
timeTics <- c(1980, 1990, 2000, 2010)
yTicsForIndexedGraphs <- c(1,2,3,4,5,6,7,8,9,10) #y tic mark locations.

maxWidth <- 6.5 #Inches
# Full page lattice plot sizes
ninePanelLayoutSpec <- c(3,3) #indicates a 3x3 arrangement of panels.
onePanelLayoutSpec <- c(1,1) #indicates a 1x1 arrangement of panels.
ninePanelGraphWidth <- maxWidth     
ninePanelGraphHeight <- 7.5  #Inches
presentationGraph1RowHeight <- 2.8 #Inches
presentationGraph2ColWidth <- 2 #Inches
presentationGraph1ColWidth <- 2*presentationGraph2ColWidth
keyTextSize <- 0.85 #85% of normal size
keyColumns <- 1 #Want only 1 column in the key for lattice graphs
defaultKeyXLoc <- 0.01 #x position of the key. This default is good for 9-panel graphs.
defaultKeyYLoc <- 0.95 #y position of the key. This default is good for 9-panel graphs.
# Parameter graphs are tricky to ensure same size. I've found that with height set to 5 in for the 
# cd graphs and 2.93 in for the sf graphs, I get a nearly-exact match with graph size. From these settings I 
# can set up 2 equations:
#    5 in = 2*hgraph + haxis
#    2.93 in = hgraph + haxis
# to estimate that h_axis = 0.86 in and hgraph = 2.05 in.
h_param_graph <- 2.07 # Inches for a single panel
h_param_axis <- 0.86 # Inches (also includes whatever margin is applied)
sfParameterGraphHeight <- h_param_graph + h_param_axis
cdParameterGraphHeight <- 2*h_param_graph + h_param_axis
cesParameterGraphHeight <- 3*h_param_graph + h_param_axis
linexParameterGraphHeight <- h_param_graph + h_param_axis
sfParameterGraphWidth <- maxWidth
cdParameterGraphWidth <- maxWidth
cesParameterGraphWidth <- maxWidth
linexParameterGraphWidth <- maxWidth
# Other graph parameters that apply to all graphs
scaleTextSize <- 0.8  #Multiple of normal size
scaleTickSize <- -0.5 #50% of normal size and pointing INWARD!
@

@@

#################
# Creates a plot with y residuals vs. energy (e). The y residuals are from the specified modelType without energy.
# modelType: one of "CD" or "CES"
# energyType: one of "Q", "X", or "U" to serve as the ordinate of the graph
# countryAbbrev: set to NA (the default) if you want a lattice graph with all countries shown. Set to one of
#                US, UK, JP, CN, ZA, SA, IR, TZ, ZM for an individual graph.
## 
createPartialResidualPlot <- function(modelType, energyType, countryAbbrev=NA, textScaling=1.0){
  xLabelString <- paste("$", tolower(energyType), "$", sep="")
  yLabelString <- paste("$y$ residuals (", modelType, " without energy)", sep="")
  if (! is.na(countryAbbrev)){
    # An individual graph for a single country is desired.
    data <- calcResiduals(modelType=modelType, countryAbbrev=countryAbbrev, energyType=energyType)
    plot <- xyplot(resid ~ iEToFit, data=data,
                   type=c("p"),
                   col="black",
                   scales=list(cex=scaleTextSize*textScaling, #controls text size on scales.
                               tck=scaleTickSize), #controls tick mark length. < 0 for inside the graph.
                   xlab=list(label=xLabelString, cex=textScaling),
                   ylab=list(label=yLabelString, cex=textScaling)
                   )
  } else {
    # Lattice graph will all countries is desired.
    factorLevels <- countryNamesAlph # We want all countries shown
    layout <- ninePanelLayoutSpec
    indexCond <- list(countryOrderForGraphs) # We want all countries in this order
    data <- do.call("rbind", lapply(countryAbbrevsAlph, calcResiduals, modelType=modelType, energyType=energyType))
    plot <- xyplot(resid ~ iEToFit | Country, data=data,
                   type = c("p"),
                   index.cond = indexCond, #orders the panels.
                   layout = layout,
                   strip = strip.custom(factor.levels=factorLevels, 
                                        bg="transparent", # Sets background transparent to match the graph itself.
                                        par.strip.text=list(cex=textScaling) # Scales text in the strip.
                                        ),
                   as.table = TRUE, #indexing of panels starts in upper left and goes across rows.
                   col = "black",
                   scales = list(cex=scaleTextSize*textScaling, #controls text size on scales.
                                 tck=scaleTickSize, #controls tick mark length. < 0 for inside the graph.
                                 alternating=FALSE), # prevents left/right, top/bot alternating
                   xlab = list(label=xLabelString, cex=textScaling),
                   ylab = list(label=yLabelString, cex=textScaling)
                   )
  }
  return(plot)
}

>>>>

# Statistical significance levels. We'll work with 95% CIs
ciLevel <- 0.95
ciHalfLevel <- ciLevel + (1.0-ciLevel)/2.0
# List of countries
countryAbbrevs <- c(US="US", UK="UK", JP="JP", CN="CN", ZA="ZA", SA="SA", IR="IR", TZ="TZ", ZM="ZM")
countryAbbrevsAlph <- c(CN="CN", IR="IR", JP="JP", SA="SA", TZ="TZ", UK="UK", US="US", ZA="ZA", ZM="ZM")
countryAbbrevsU <- c(US="US", UK="UK", JP="JP") #Only these countries have useful work data
countryAbbrevsAlphU <- c(JP="JP", UK="UK", US="US")
countryNamesAlph <- c(CN="China", IR="Iran", JP="Japan", SA="Saudi Arabia", TZ="Tanzania", UK="United Kingdom", US="USA", ZA="South Africa", ZM="Zambia") #In alphabetical order.
yLimitsForGDPGraphs <- list(c(1,10), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4)) # Alph order
energyTypes <- c(Q="Q", X="X", U="U") # List of energy types
factors <- c(K="K", L="L", Q="Q", X="X", U="U") # List of factors of production
########### Several global parameters for graphs. Set here and use below to ensure consistent appearance of graphs.
# Set the order for presenting countries in 3x3 lattice graphs. Default is alphabetical. 
# "1" means first alphabetically.
countryOrderForGraphs <- c(7,6,3,1,8,5,4,2,9) # Sets the order as US, UK, JP, CN, ZA, TZ, SA, IR, ZM.
countryOrderForGraphsU <- c(3, 2, 1) # Sets the order as US, UK, JP when we're looking only at energyType=U.
timeTics <- c(1980, 1990, 2000, 2010)
yTicsForIndexedGraphs <- c(1,2,3,4,5,6,7,8,9,10) #y tic mark locations.

maxWidth <- 6.5 #Inches
# Full page lattice plot sizes
ninePanelLayoutSpec <- c(3,3) #indicates a 3x3 arrangement of panels.
threePanelLayoutSpec <- c(3,1) #indicates a 1 row x 3 column arangement of panels
onePanelLayoutSpec <- c(1,1) #indicates a 1x1 arrangement of panels.
ninePanelGraphWidth <- maxWidth     
ninePanelGraphHeight <- 7.5  #Inches
presentationGraph1RowHeight <- 2.8 #Inches
presentationGraph2ColWidth <- 2 #Inches
presentationGraph1ColWidth <- 2*presentationGraph2ColWidth
keyTextSize <- 0.85 #85% of normal size
keyColumns <- 1 #Want only 1 column in the key for lattice graphs
defaultKeyXLoc <- 0.01 #x position of the key. This default is good for 9-panel graphs.
defaultKeyYLoc <- 0.95 #y position of the key. This default is good for 9-panel graphs.
# Parameter graphs are tricky to ensure same size. I've found that with height set to 5 in for the 
# cd graphs and 2.93 in for the sf graphs, I get a nearly-exact match with graph size. From these settings I 
# can set up 2 equations:
#    5 in = 2*hgraph + haxis
#    2.93 in = hgraph + haxis
# to estimate that h_axis = 0.86 in and hgraph = 2.05 in.
h_param_graph <- 2.07 # Inches for a single panel
h_param_axis <- 0.86 # Inches (also includes whatever margin is applied)
sfParameterGraphHeight <- h_param_graph + h_param_axis
cdParameterGraphHeight <- 2*h_param_graph + h_param_axis
cesParameterGraphHeight <- 3*h_param_graph + h_param_axis
linexParameterGraphHeight <- h_param_graph + h_param_axis
sfParameterGraphWidth <- maxWidth
cdParameterGraphWidth <- maxWidth
cesParameterGraphWidth <- maxWidth
linexParameterGraphWidth <- maxWidth
# Other graph parameters that apply to all graphs
scaleTextSize <- 0.8  #Multiple of normal size
scaleTickSize <- -0.5 #50% of normal size and pointing INWARD!
@

@@

#################
# Creates a plot with y residuals vs. energy (e). The y residuals are from the specified modelType without energy.
# modelType: one of "CD" or "CES"
# energyType: one of "Q", "X", or "U" to serve as the ordinate of the graph
# countryAbbrev: set to NA (the default) if you want a lattice graph with all countries shown. Set to one of
#                US, UK, JP, CN, ZA, SA, IR, TZ, ZM for an individual graph.
## 
createPartialResidualPlot <- function(modelType, energyType, countryAbbrev=NA, textScaling=1.0){
  xLabelString <- paste("$", tolower(energyType), "$", sep="")
  yLabelString <- paste("$y$ residuals (", modelType, " without energy)", sep="")
  if (! is.na(countryAbbrev)){
    # An individual graph for a single country is desired.
    data <- calcResiduals(modelType=modelType, countryAbbrev=countryAbbrev, energyType=energyType)
    plot <- xyplot(resid ~ iEToFit, data=data,
                   type=c("p"),
                   col="black",
                   scales=list(cex=scaleTextSize*textScaling, #controls text size on scales.
                               tck=scaleTickSize), #controls tick mark length. < 0 for inside the graph.
                   xlab=list(label=xLabelString, cex=textScaling),
                   ylab=list(label=yLabelString, cex=textScaling)
                   )
  } else {
    # Lattice graph will all countries is desired.
    data <- do.call("rbind", lapply(countryAbbrevsAlph, calcResiduals, modelType=modelType, energyType=energyType))
    # If useful work is desired, need to trim the data set.
    if (energyType == "U"){
      # Use data from only those countries that have U defined.
      data <- subset(data, !is.na(iEToFit))
print(data)
      layout <- threePanelLayoutSpec
      factorLevels <- countryAbbrevsAlphU # Want only those countries for which U is known
print(factorLevels)
      indexCond <- list(countryOrderForGraphsU) # Orders as US, UK, JP
    } else {
      # Use data from all countries, because all countries have Q and X data.
      factorLevels <- countryNamesAlph # We want all countries shown
      layout <- ninePanelLayoutSpec
      indexCond <- list(countryOrderForGraphs) # Orders as US, UK, JP, CN, ZA, SA, IR, TZ, ZM
    }
    plot <- xyplot(resid ~ iEToFit | Country, data=data,
                   type = c("p"),
                   index.cond = indexCond, #orders the panels.
                   layout = layout,
                   strip = strip.custom(factor.levels=factorLevels, 
                                        bg="transparent", # Sets background transparent to match the graph itself.
                                        par.strip.text=list(cex=textScaling) # Scales text in the strip.
                                        ),
                   as.table = TRUE, #indexing of panels starts in upper left and goes across rows.
                   col = "black",
                   scales = list(cex=scaleTextSize*textScaling, #controls text size on scales.
                                 tck=scaleTickSize, #controls tick mark length. < 0 for inside the graph.
                                 alternating=FALSE), # prevents left/right, top/bot alternating
                   xlab = list(label=xLabelString, cex=textScaling),
                   ylab = list(label=yLabelString, cex=textScaling)
                   )
  }
  return(plot)
}

### 2013-04-30 EconGrowth.Rnw
Partial residual plots for U now include only US, UK, and JP, the countries for which U data is available.

\item{Heun: Add references from Lucas}

@@

\item{de Wit: provide funding acknowledgement, if necessary.}

@@

@@

@@

In this paper, we assess three prominent energy-augmented production functions appear in the literature: Cobb-Douglas (CDe), Constant Elasticity of Substitution (CESe), and LINear EXponential (LINEX). The following sections describe each.

@@

\begin{table} \caption{Food end use to intake ratios for all countries.} \label{tab:f_enduse_intake}
  \begin{center}
    \begin{tabular}{r c c c c c c c c c}
      Country & US & UK & JP & CN & ZA & SA & IR & TZ & ZM \\
      \hline
      $f_{enduse/intake}$ & 0.62 & 0.65 & 0.76 & 0.76 & 0.87 & 0.72 & 0.72 & 0.72 & 0.72\\
    \end{tabular}
  \end{center}
\end{table}

@@

<<CDPartialResidualsU, fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.align="center", fig.cap="Cobb-Douglas partial residuals plot for useful work ($u$).", fig.pos="H">>=
createPartialResidualPlot(modelType="CD", energyType="U", countryAbbrev=NA)
@

\section{Cobb-Douglas Models With Energy}

@@

\subsection{CES without energy (CES)}

@@

\subsection{CES with Thermal Energy}

@@

\subsection{CES with Exergy}

@@

\subsection{CES with Useful Work}

@@

\subsection{CES comparisons}

@@

\subsection{LINEX with Thermal Energy}

@@

\subsection{LINEX with Exergy}

@@

\subsection{LINEX with Useful Work}

@@

\subsection{LINEX comparisons}

@@

Funding for Caleb Reese and Lucas Timmer was generously supplied by the Jack and Lois Kuipers Applied Mathematics Endowment and the Calvin College Alumni Association. Support for Matthew Kuperus Heun was provided by a Calvin College Research Followship. Funding for Martin de Wit was provided by *******. The authors thank Loren L. Heun (Western Michigan University) for insightful critique of the statistical analyses presented herein. The authors are grateful to Benjamin S. Warr (INSEAD) for providing the REXS database \citep{Warr:2012a}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Model Parameters}
\setcounter{table}{0} % Restarts table number counter at 0 in this section.
\setcounter{figure}{0} % Restarts figure number counter at 0 in this section.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Single-factor Models}

@@

\subsection{Cobb-Douglas Models}

\subsection{CES Models}

\subsection{LINEX Models}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Goodness of fit}
\setcounter{table}{0} % Restarts table number counter at 0 in this section.
\setcounter{figure}{0} % Restarts figure number counter at 0 in this section.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

@@

@@

\item{\st{Heun: Add references from Lucas}}

@@

\item{\st{de Wit: provide funding acknowledgement, if necessary}}

@@

In this paper, we assess three prominent energy-augmented production functions appear in the literature: Cobb-Douglas (CDe), Constant Elasticity of Substitution (CESe), and LINear EXponential (LINEX). The following subsections describe each.

@@

\begin{table} \caption{Food enduse-to-intake ratios for all countries.} \label{tab:f_enduse_intake}
  \begin{center}
    \begin{tabular}{r c c c c c c c c c}
      Country & US & UK & JP & CN & ZA & SA & IR & TZ & ZM \\
      \hline
      $f_{enduse/intake}$ & 0.62 & 0.65 & 0.76 & 0.76 & 0.87 & 0.72 & 0.72 & 0.72 & 0.72\\
    \end{tabular}
  \end{center}
\end{table}

@@

<<CDPartialResidualsU, fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.align="center", fig.cap="Cobb-Douglas partial residuals plot for useful work ($u$).", fig.pos="H", eval=TRUE>>=
createPartialResidualPlot(modelType="CD", energyType="U", countryAbbrev=NA)
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Cobb-Douglas Models With Energy}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%++++++++++++++++++++++++++++++
\subsection{CES without energy (CES)}
%++++++++++++++++++++++++++++++

@@

%++++++++++++++++++++++++++++++
\subsection{CES with Thermal Energy}
%++++++++++++++++++++++++++++++

@@

%++++++++++++++++++++++++++++++
\subsection{CES with Exergy}
%++++++++++++++++++++++++++++++

@@

%++++++++++++++++++++++++++++++
\subsection{CES with Useful Work}
%++++++++++++++++++++++++++++++

@@

%++++++++++++++++++++++++++++++
\subsection{CES comparisons}
%++++++++++++++++++++++++++++++

@@

%++++++++++++++++++++++++++++++
\subsection{LINEX with Thermal Energy}
%++++++++++++++++++++++++++++++

@@

%++++++++++++++++++++++++++++++
\subsection{LINEX with Exergy}
%++++++++++++++++++++++++++++++

@@

%++++++++++++++++++++++++++++++
\subsection{LINEX with Useful Work}
%++++++++++++++++++++++++++++++

@@

%++++++++++++++++++++++++++++++
\subsection{LINEX comparisons}
%++++++++++++++++++++++++++++++

@@

Funding for Caleb Reese and Lucas Timmer was generously supplied by the Jack and Lois Kuipers Applied Mathematics Endowment and the Calvin College Alumni Association. Support for Matthew Kuperus Heun was provided by a Calvin College Research Followship. Partial funding for Martin de Wit was provided by the South African National Research Foundation (NRF).\footnote{Any opinion, findings, and conclusions or recommendations expressed herein are those of the authors, and, therefore, NRF does not accept any liability in regard thereto.} The authors are grateful to Benjamin S. Warr (INSEAD) for providing the REXS database \citep{Warr:2012a}. The authors thank Loren L. Heun (Western Michigan University) for insightful critique of the statistical analyses presented herein.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Model Parameters}
\label{sec:Model_Parameters}
\setcounter{table}{0} % Restarts table number counter at 0 in this section.
\setcounter{figure}{0} % Restarts figure number counter at 0 in this section.
\setcounter{equation}{0} % Restarts equation counter to 0 in this section.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%++++++++++++++++++++++++++++++
\subsection{Single-factor Models}
%++++++++++++++++++++++++++++++

****** This subsection should be either (a) deleted or (b) moved into the main body of the paper. I would prefer to delete it on the grounds that single-factor models are not often used in the literature. However, I want to talk with Martin about this before making the decision.  So, I'm parking it here for now. --Matt. ********

%++++++++++++++++++++++++++++++
\subsection{Cobb-Douglas Models}
%++++++++++++++++++++++++++++++

%++++++++++++++++++++++++++++++
\subsection{CES Models}
%++++++++++++++++++++++++++++++

%++++++++++++++++++++++++++++++
\subsection{LINEX Models}
%++++++++++++++++++++++++++++++

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Goodness of fit}
\label{sec:Goodness_of_Fit}
\setcounter{table}{0} % Restarts table number counter at 0 in this section.
\setcounter{figure}{0} % Restarts figure number counter at 0 in this section.
\setcounter{equation}{0} % Restarts equation counter to 0 in this section.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Single-factor Models}
\label{sec:Single-factor_Models}
\setcounter{table}{0} % Restarts table number counter at 0 in this section.
\setcounter{figure}{0} % Restarts figure number counter at 0 in this section.
\setcounter{equation}{0} % Restarts equation counter to 0 in this section.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

****** This section should be either (a) deleted or (b) moved into the main body of the paper. I would prefer to delete it on the grounds that single-factor models are not often used in the literature. However, I want to talk with Martin about this before making the decision.  So, I'm parking it here for now. --Matt. ********

@@
